### 2016-04-27 Paper_v5.tex
Polish and refine

Cells process nutrients and respond to changes in their environment using complex enzyme catalyzed biochemical networks.
Mathematical modeling has evolved as a powerful paradigm to analyze, and ultimately design these complex networks \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}. Mathematical modeling of biochemical networks is often an iterative process.
First, models are formulated from existing biochemical knowledge, and then model parameters are estimated using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
Parameter estimation is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental measurements and model simulations is minimized using an optimization strategy \cite{moles2003parameter}.
Optimal parameter estimates obtained from model training are then used to validate the model on unseen experimental data.
If the validation studies fail, model construction and calibration are repeated iteratively until satisfactory results are obtained.

Parameter estimation is a major challenge in the iterative development of biochemical models.
Although parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
the complex dynamics of large biological systems and noisy, often incomplete experimental data sets pose a unique challenge.
Often optimization problems involving biological systems are non-linear and multi-modal i.e., typical models often have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}.
Non-linearity coupled with multi-modality generally renders local optimization techniques such as pattern search \cite{hooke1961direct}, Nelder-Mead simplex methods \cite{nelder1965simplex}, steepest descent methods or Levenberg-Marquardt \cite{more1978levenberg} incapable of reliably obtaining globally optimal solutions as these methods often terminate at local minimum.
Though deterministic global optimization techniques (for example algorithms based on branch and bound) can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global},
the absence of derivative information, the discontinuity of the objective functions, non-smooth regions or the lack of knowledge about the objective function hampers these techniques.

@@

Meta-heuristic optimization approaches like Genetic Algorithms (GAs) \cite{goldberg2006genetic}, Simulated Annealing (SA) \cite{kirkpatrick1983optimization},
Evolutionary Programming \cite{fogel2009artificial} and Differential Evolution (DE) \cite{storn1997differential,tsai2005evolutionary,wang2001hybrid,noman2007inferring} have shown promise on non-linear multi-modal problems \cite{sun2012parameter}.
These techniques do not make any assumptions, nor do they require, \textit{a~priori} information about the structure of objective function.
Meta-heuristics are often very effective at finding globally optimal or near optimal solutions.
For example, Mendes et al. used simulated annealing to estimate rate constants for the irreversible inhibition of HIV proteinase \cite{mendes1998non},
while Modchang et al. used a genetic algorithm (GA) to estimate parameters for a model of G-protein-coupled receptors (GPCRs) activity \cite{modchang2008mathematical}. Parameter estimates obtained using GA helped qualitatively determine the effectiveness of two G-protein agonists, N6-cyclopentyladenosine (CPA) and 5'-N-ethylcarboxamidoadenosine (NECA).
Tashkova et al. compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis
and showed that DE was the most effective of the approaches tested \cite{tashkova2011parameter}.
Banga and co-workers have also successfully applied scatter-search methods to estimate parameters on non-linear biological processes \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter}.
Hybrid approaches which combine meta-heuristics with local optimization techniques have also become popular.
For example, Villaverde et al. developed the enhanced scatter search (eSS) method, which combined scatter search with local search methods,
for parameter estimation in a collection of systems biology models \cite{villaverde2015biopredyn}.
However, despite these successes, a major drawback of most meta-heuristics is the large number of function evaluations required to explore parameter space.
Performing a large number of potentially expensive function evaluations is not desirable (and perhaps not feasible) for many types of biochemical models.
Alternatively, Tolson and Shoemaker found, using high-dimensional watershed models, that perturbing only a subset of parameters was an effective strategy for estimating
parameters in models with expensive function evaluations \cite{tolson2007dynamically}. This approach called Dynamically Dimensioned Search (DDS) is a simple stochastic single-solution based heuristic that estimates nearly optimal solutions within a specified maximum function (or model) evaluation limit.
Taken together, meta-heuristics are often very effective at finding globally optimal or near optimal solutions, however the large number of function evaluations required to
converge to a solution limits their application for expensive biochemical models.

@@

In this study, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel meta-heuristic that combines the global search capability of multi-swarm particle swarm optimization with the greedy refinement of dynamically dimensioned search (DDS).
The objective of DOPS is to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
DOPS uses multi-swarm particle swarm optimization to generate candidate solutions, which are then greedily updated using dynamically dimensioned search.
First, we tested the performance of DOPS on commonly used test functions for global optimization (Ackley and Rosenbrock) and published biochemical benchmark problems. Next, we used DOPS to identify a model of human coagulation cascade.
DOPS outperformed common meta-heuristic approaches like differential evolution, simulated annealing, and dynamically dimensioned search on the test functions, benchmark problems and the coagulation model. For example, DOPS recovered the nominal parameters for the benchmark problems using an order of magnitude less function evaluations than eSS in all cases. Taken together, these studies suggest DOPS is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.

@@

\subsection*{DOPS explores parameter space using a combination of global and greedy search.}
DOPS is a novel meta-heuristic which combines multi-swarm particle swarm methods with the dynamically dimensioned search approach of Shoemaker and colleagues (Fig. \ref{fig-algorithm}).
The goal of DOPS is to estimate optimal or near optimal parameter vectors for high-dimensional biological models within a specified number of function evaluations.
Toward this objective, DOPS begins by using a particle swarm search and then dynamically switches, using an adaptive switching criteria, to the DDS search phase.
The particle swarm search uses multiple sub-swarms wherein each solution represented by a particle is updated using equation \ref{eqn:update-rule}. The particle update is influenced by the best particle amongst all particles in the sub-swarm and the best solution found by the particle till the current iteration. This update rule however differs from a conventional PSO update wherein there are no velocity terms involved. The particle updates continue to happen within the sub-swarms for a certain number of iterations (or function evaluations) after which the sub-swarms are reorganized which is similar to the regrouping strategy described by Zhao et al. \cite{zhao2008dynamic}. DOPS switches out of this PSO phase based on an adaptive switching criteria that is a function of error convergence rate. If the error represented by the best particle does not drop for a threshold number of function evaluations, DOPS switches automatically to the DDS search phase. The DDS search uses the best particle from the PSO search phase as its initial solution or candidate vector. Thereafter, the particle is greedily updated by perturbing a subset of dimensions for the remaining number of function evaluations. The number of dimensions perturbed is a monotonically decreasing function that generally depends on the number of function evaluations within the DDS phase.

\subsection*{DOPS minimized benchmark problems with a fixed number of function evaluations.}
We tested DOPS and four other meta-heuristic approaches on the minimization of the Ackley and Rastrigin functions.
DOPS performed similarly or outperformed the other meta-heuristic approaches on both of these test functions (Fig.\ref{fig-testfunctions}).
The Ackley and Rastrigin functions both have multiple local extrema and attain a global minimum value of zero.
In each case, we fixed the maximum number of function evaluations at $\mathcal{N}$ = 4000 and ran 25 independent experiments with different initial parameter vectors.
DOPS found optimal or near optimal solutions for both the 10-dimensional Ackley (Fig. \ref{fig-testfunctions}A) and Rastrigin (Fig. \ref{fig-testfunctions}B) functions within the budget of function evaluations.
In each of the 10-dimensional cases, other meta-heurtistics such as DDS and DE also performed well, however DOPS consistently outperformed all other approaches tested.
This performance difference was more pronounced as the dimension of the search problem increased; for a 300-dimensional Rastrigin function DOPS was the only approach to find an optimal or near optimal solution
within the function evaluation budget (Fig. \ref{fig-testfunctions}B).
Taken together, DOPS performed at least as well as other meta-heuristic approaches on small test problems, but was especially suited to large dimension search spaces.
Next, we tested DOPS on benchmark biochemical models of varying complexity.

Villaverde and co-workers published a set of benchmark biochemical problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
They ranked ordered these example problems by computational cost from most to least expensive.
We evaluated the performance of DOPS on problems from the least and most expensive categories.
The least expensive problem (henceforth referred to as problem B4) was a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters \cite{villaverde2014high}.
The biochemical reactions were modeled using modular rate laws and the generalized form of Michaelis-Menten kinetics.
The expensive problem was a genome scale kinetic model of \textit{Saccharomyces~cerevisiae} with 261 reactions, 262 variables and 1759 parameters \cite {smallbone2013large}(henceforth referred to as problem B1). In both cases, synthetic time series data sets were generated using known parameter values, these data were then used by the parameter estimation methods.
For problem B1, the time series data consisted of 44 observables, and for problem B4 the data corresponded to 13 different metabolite measurement sets.
We fixed the number of function evaluations at $\mathcal{N}$ = 4000, and trained both models against the synthetic experimental data.
In both cases we found good fits  (Fig. \ref{fig-sims-b1} and Fig. \ref{fig-sims-b4}) to the problems.
We also recaptured the nominal parameter values within the specified number of function evaluations (Fig. \ref{fig-benchmark}).
Thus, DOPS estimated the parameters in benchmark biochemical models, and recovered the original parameters from synthetic data.
Next, we compared the performance of DOPS with the four other meta-heuristic approaches for a model of the human coagulation cascade.

>>>>

Cells process nutrients and respond to changes in their environment using complex enzyme catalyzed biochemical networks.
Mathematical modeling has evolved as a powerful paradigm to analyze, and ultimately design these complex networks \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}. Mathematical modeling of biochemical networks is often an iterative process.
First, models are formulated from existing biochemical knowledge, and then model parameters are estimated using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
Parameter estimation is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental measurements and model simulations is minimized using an optimization strategy \cite{moles2003parameter}.
Optimal parameter estimates are then used to predict unseen experimental data.
If the validation studies fail, model construction and calibration are repeated iteratively until satisfactory results are obtained.
As our biological knowledge increases, model formulation may not be as significant a challenge, but parameter estimation will likely remain difficult.

Parameter estimation is a major challenge to the development of biochemical models.
Parameter estimation has been a well studied engineering problem for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse}.
However, the complex dynamics of large biological systems and noisy, often incomplete experimental data sets pose a unique estimation challenge.
Often optimization problems involving biological systems are non-linear and multi-modal i.e., typical models have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}.
Non-linearity coupled with multi-modality renders local optimization techniques such as pattern search \cite{hooke1961direct}, Nelder-Mead simplex methods \cite{nelder1965simplex}, steepest descent or Levenberg-Marquardt \cite{more1978levenberg} incapable of reliably obtaining globally optimal solutions as these methods often terminate at local minimum.
Though deterministic global optimization techniques (for example algorithms based on branch and bound) can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global},
the absence of derivative information, discontinuous objective functions, non-smooth regions or the lack of knowledge about the objective function hampers these techniques.

@@

Meta-heuristics like Genetic Algorithms (GAs) \cite{goldberg2006genetic}, Simulated Annealing (SA) \cite{kirkpatrick1983optimization},
Evolutionary Programming \cite{fogel2009artificial} and Differential Evolution (DE) \cite{storn1997differential,tsai2005evolutionary,wang2001hybrid,noman2007inferring} have all shown promise on non-linear multi-modal problems \cite{sun2012parameter}.
These techniques do not make any assumptions, nor do they require, \textit{a~priori} information about the structure of the objective function.
Meta-heuristics are often very effective at finding globally optimal or near optimal solutions.
For example, Mendes et al. used SA to estimate rate constants for the inhibition of HIV proteinase \cite{mendes1998non},
while Modchang et al. used a GA to estimate parameters for a model of G-protein-coupled receptor (GPCR) activity \cite{modchang2008mathematical}.
Parameter estimates obtained using the GA stratified the effectiveness of two G-protein agonists, N6-cyclopentyladenosine (CPA) and 5'-N-ethylcarboxamidoadenosine (NECA).
Tashkova et al. compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis;
DE was the most effective of the approaches tested \cite{tashkova2011parameter}.
Banga and co-workers have also successfully applied scatter-search to estimate model parameters \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter}.
Hybrid approaches, which combine meta-heuristics with local optimization techniques, have also become popular.
For example, Villaverde et al. developed the enhanced scatter search (eSS) method, which combined scatter and local search methods,
for parameter estimation in biological models \cite{villaverde2015biopredyn}.
However, despite these successes, a major drawback of most meta-heuristics remains the large number of function evaluations required to explore parameter space.
Performing numerous potentially expensive function evaluations is not desirable (and perhaps not feasible) for many types of biochemical models.
Alternatively, Tolson and Shoemaker found, using high-dimensional watershed models, that perturbing only a subset of parameters was an effective strategy for estimating
parameters in expensive models \cite{tolson2007dynamically}. Their approach, called Dynamically Dimensioned Search (DDS),
is a simple stochastic single-solution heuristic that estimates nearly optimal solutions within a specified maximum number of function (or model) evaluations.
Thus, while meta-heuristics are often effective at estimating globally optimal or nearly optimal solutions, they require a large number of function evaluations to
converge to a solution.

@@

In this study, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel hybrid meta-heuristic that combines the global search capability of multi-swarm particle swarm optimization with the greedy refinement of dynamically dimensioned search (DDS).
The objective of DOPS is to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
DOPS uses multi-swarm particle swarm optimization to generate nearly optimal candidate solutions, which are then greedily updated using dynamically dimensioned search.
We tested DOPS using a combination of classic optimization test functions, biochemical benchmark problems and real-world biochemical models.
First, we tested the performance of DOPS on commonly used test functions for global optimization (Ackley and Rosenbrock) and published biochemical benchmark problems.
Next, we used DOPS to estimate the parameters of a model of the human coagulation cascade.
On average, DOPS outperformed other common meta-heuristics like differential evolution, simulated annealing, single-swarm particle swarm optimization, and dynamically dimensioned search on the optimization test functions, benchmark problems and the coagulation model. For example, DOPS recovered the nominal parameters for the benchmark problems using an order of magnitude fewer function evaluations than eSS in all cases.
It also produced parameter estimates for the coagulation model that predicted unseen coagulation data sets.
Thus, DOPS is a promising hybrid meta-heuristic for the estimation of biochemical model parameters in relatively few function evaluations.

@@

\subsection*{DOPS explores parameter space using a combination of global methods.}
DOPS is a novel hybrid meta-heuristic which combines a multi-swarm particle swarm method with the dynamically dimensioned search approach of Shoemaker and colleagues (Fig. \ref{fig-algorithm}).
The goal of DOPS is to estimate optimal or near optimal parameter vectors for high-dimensional biological models within a specified number of function evaluations.
Toward this objective, DOPS begins by using a multi-swarm particle swarm search and then dynamically switches, using an adaptive switching criteria, to the DDS approach.
The particle swarm search uses multiple sub-swarms wherein the update to each particle (corresponding to a parameter vector estimate) is influenced by the best particle amongst the sub-swarm, and the current globally best particle.
Particle updates occur within sub-swarms for a certain number of function evaluations, after which the sub-swarms are reorganized.
This sub-swarm mixing is similar to the regrouping strategy described by Zhao et al. \cite{zhao2008dynamic}. DOPS switches out of the particle swarm phase based upon an adaptive switching criteria that is a function of the rate of error convergence.
If the error represented by the best particle does decrease for a threshold number of function evaluations, DOPS switches automatically to the DDS search phase.
The DDS search is initialized with the globally best particle from the particle swarm phase, thereafter, the particle is greedily updated by perturbing a subset of dimensions for the remaining number of function evaluations.
The identity of the parameters perturbed is chosen randomly, with fewer parameters perturbed the higher the number of function evaluations.

%We tested DOPS and four other meta-heuristic approaches on the minimization of the Ackley and Rastrigin functions.

\subsection*{DOPS minimized benchmark problems using fewer function evaluations.}
On average, DOPS performed similarly or outperformed the four other meta-heuristics for the Ackley and Rastrigin test functions (Fig. \ref{fig-testfunctions}).
The Ackley and Rastrigin functions both have multiple local extrema and attain a global minimum value of zero.
In each case, we fixed the maximum number of function evaluations at $\mathcal{N}$ = 4000 and ran $\mathcal{T}$ = 25 independent experiments with different initial parameter vectors.
DOPS found optimal or near optimal solutions for both the 10-dimensional Ackley (Fig. \ref{fig-testfunctions}A) and Rastrigin (Fig. \ref{fig-testfunctions}B) functions within the budget of function evaluations.
In each of the 10-dimensional cases, other meta-heurtistics such as DDS and DE also performed well. However, DOPS consistently outperformed all other approaches tested.
This performance difference was more pronounced as the dimension of the search problem increased; for a 300-dimensional Rastrigin function, DOPS was the only approach to find an optimal or near optimal solution
within the function evaluation budget (Fig. \ref{fig-testfunctions}B).
Taken together, DOPS performed at least as well as other meta-heuristic approaches on small dimensional test problems, but seemed especially suited to large dimensional search spaces.
Next, we tested DOPS on benchmark biochemical models of varying complexity.

Villaverde and co-workers published a set of benchmark biochemical problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}.
They ranked the example problems by computational cost from most to least expensive.
We evaluated the performance of DOPS on problems from the least and most expensive categories.
The least expensive problem (henceforth referred to as problem B4) was a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters \cite{villaverde2014high}.
The biochemical reactions were modeled using modular rate laws and generalized Michaelis-Menten kinetics.
On the other hand, the expensive problem was a genome scale kinetic model of \textit{Saccharomyces~cerevisiae} with 261 reactions, 262 variables and 1759 parameters \cite{smallbone2013large} (henceforth referred to as problem B1).
In both cases, synthetic time series data sets, generated using known parameter values, were used to estimate model parameters.
For problem B1, the time series data consisted of 44 observables, and for problem B4 the data corresponded to 13 different metabolite measurement sets.
We fixed the number of function evaluations at $\mathcal{N}$ = 4000, and trained both models against the synthetic experimental data.
In both cases DOPS, produced good fits to the synthetic data (Fig. \ref{fig-sims-b1} and Fig. \ref{fig-sims-b4}), and recapitulated the nominal parameter values using only $\mathcal{N}\leq~4000$ function evaluations (Fig. \ref{fig-benchmark}).
On the other hand, enhanced scatter search (eSS) with a local optimizer took on order $10^{5}$ function evaluations for the same problems.
Thus, DOPS estimated the parameters in benchmark biochemical models, and recovered the original parameters from synthetic data, using fewer function evaluations.
Next, we compared the performance of DOPS with the four other meta-heuristics for a model of the human coagulation cascade.
